Predicting Conflict Escalation

### 2023-05-04 docs/hes_final/article.qmd
More changes to results

We evaluated the models against a set of common performance metrics (see @tbl-best-max-metrics, @tbl-best-min-metrics, @fig-roc, @fig-prc). Despite conceptual differences in dataset creation and modeling techniques, all three of the described above approaches produced comparative results. Each of the models outperformed the other two in a subset of these metrics. This suggests that the choice of the model architecture for this task would depend on specific application goals. We assume that UNICEF decision makers would appreciate a tool that flags all potential rising conflicts even if sometimes it has false positives. At the same time, if the false positive rate is too high the signal would become too noisy for the tool to be useful. Given these assumptions and the class imbalance of the original dataset, Recall (ratio of correctly identified conflict spikes to the total number of true spikes) and Cohen Kappa Score (overall agreement between predictions and true labels adjusted for class imbalance) seem to be most relevant metrics. Two models using country features seem to perform better on these two metrics.

In this work we defined the target in a way that was not used in prior literature, we cannot directly compare our models' performance to others. The most similar work used a presense or absense of any armed conflict related fatality as their binary target  [@10.1093/jeea/jvac025]. They focused on ROC_AUC as their metric and were able to achieve 0.83 on overall dataset and 0.75 on hard cases (sudden conflict in peaceful countries) with their text based (LDA topic retrieval) model. Our best ROC_AUC score is 69% but the comparison is not very meaningful due to difference in targets.

>>>>

We evaluated the models against a set of common performance metrics (see @tbl-best-max-metrics, @tbl-best-min-metrics, @fig-roc, 
@fig-prc). Despite conceptual differences in dataset creation and modeling techniques, all three of the described above approaches 
produced comparative results. Each of the models outperformed the other two in a subset of these metrics.

Our target was defined differently from related work which made direct comparison less meaningful.
The closest approach used a presense or absense of any armed conflict related fatality as their binary target  [@10.1093/jeea/jvac025]. 
They focused on ROC_AUC as their metric and were able to achieve 0.83 on overall dataset and 0.75 on hard cases (sudden conflict in 
peaceful countries) with their text based (LDA topic extraction) model. Our best ROC_AUC score is 69% (achieved by text + country labels 
XGBoost/KNN model).

We assume that UNICEF decision makers would appreciate a tool that flags all potential rising conflicts even if sometimes it has false 
positives. At the same time, if the false positive rate is too high the signal would become too noisy for the tool to be useful. Given 
these assumptions and the class imbalance of the original dataset, Recall (ratio of correctly identified conflict spikes to the total 
number of true spikes) and Cohen Kappa Score (overall agreement between predictions and true labels adjusted for class imbalance) seem 
to be most relevant metrics. The relative ranks of the models are the opposite for these two metrics with each of the two models
with additional country data ranking best on one and worst on the other and the text-only model always being in the middle.
